---
title: 'Module 3: Matrices, Iterating, Applying, Deciding, and Lists'
author: "Ted Laderas"
date: "August 2, 2015"
output: html_document
---

Remember to change your working directory!

Load up the module workspace
```{r}
load("module3.RData")
```

# Part 1: Iterating

- Doing something with each weight
- Using `for` loops

## `for` loop construction on a vector

Let's calculate the total sum of the vector. Initialize a variable `sumweights`
to zero.

```{r}
sumweights <- 0
for(i in 1:length(weights)){
    sumweights <- sumweights + weights[i]
}
sumweights
```

## Alternate `for`-loop construction

You can use this alternate `for`-loop when you don't need to access the index.
This is very similar to python.

```{r}
sumweights <- 0
for(we in weights){
    sumweights <- sumweights + we
}
sumweights
```

**QUESTION 1-1**: How do we get the average using `sumweights`? Hint: What would
be a programmatic way to get `n`?
```{r}
# space for your answer here

```

`while()` loops are also available when you have a condition. Note that `cat()`
is like `print` in other languages, in that it concatenates the objects you pass
to it.

```{r}
sumweights <- 0
while(sumweights < 100){
    sumweights <- sumweights + 5
    cat(sumweights, "\n")
}
sumweights
```

When we talk about functions, we can talk more about a more programmatic way to
process elements other than a `for` loop.

# Part 2: Matrices
(for more on matrices, please refer to "Introduction to R" Chapter 5)

Let's instantiate a matrix:
Note that when you pass a vector of data to the `matrix()` function, you have it
fill by row (i.e. fills from left to right until a row is full, then starts a
tthe next row) or by column (i.e. fills until a colum is full, then starts at
the top of the next column). Note that to fill by column, you set `byrow=FALSE`.

```{r}
mat <- matrix(ncol=3, nrow=4, byrow=TRUE, data=c(1:12))
mat
```

Matrices are like vectors in they must all be the same data type (e.g.
character, numeric, boolean, etc) but more of what you learned for data frames
can also be used on matrices. We can add new row names and column names to the
matrix. 

```{r}
rownames(mat) <- c("EH", "BE", "SI", "DI")
rownames(mat)
colnames(mat) <- c("Q", "R", "S")
colnames(mat)
mat
```

Other properties of matrices are the same as data frames.

```{r}
# shows the number of rows
nrow(mat)

# shows the number of columns
ncol(mat)
```

Accessing rows and columns is similar to data frames in which we can call a
column either by number of by name.

```{r}
mat[,"Q"]
mat[,1]
```

Note, however, unlike data frames, `mat$Q` doesn't work. If you need
`data.frame` accessors, you can always cast a matrix to a `data.frame`:

```{r}
dfMat <- data.frame(mat)
dfMat$Q
```

Similarly, we can access rows.

```{r}
mat["BE",]
mat[2,]
```

**QUESTION 2-1**: For each colum of `mat`, print the name and the column sum.
Hint: iterate from 1 to `ncol(mat)`.
```{r}
# space for your answer here

```

**QUESTION 2-2**: How do we access multiple rows or columns? Grab the first two
rows of `mat`.
```{r}
# space for your answer here

```

Accessing a particular cell of a matrix:
```{r}
mat[1,2]
```

We can mix and match:
```{r}
mat["EH", 3]
```

We can manipulate the cells by:

**direct assignment**
```{r}
mat[1,2] <- 33
mat
```
**arithmetic**
```{r}
mat[1,1] <- mat[1,1] * 5
mat
```

# Part 3: If/conditional statements
- If statements

In general, `if` statements work much like any other language. Here's an `if`
statement in a `for` loop:

```{r}
for(i in 1:length(weights)){
    if(weights[i] < 50) {
        cat(weights[i], "\n")
    }
}
```

Conditional statements can be nested or chained.

```{r}
for(i in 1:length(weights)){
    if(weights[i] < 50){
        if(weights[i] > 20){
            cat(weights[i], "\n")
        }
    }
}
```

Use of the `&&` operator to chain conditional statements with `AND`. `OR` can be
specified with `||`. Note that `&&` and `||` work with single value
conditionals, while `&` and `|` work with vectors (as we have seen for
filtering.

```{r}
for(i in 1:length(weights)){
    if(weights[i] < 50 && weights[i] > 20){
        cat(weights[i], "\n")
    }
}
```

`else` is also an option.
```{r}
for(i in 1:length(weights)){
    if(weights[i] < 50 && weights[i] > 20){
        cat(weights[i], "\n")
    }
    else{cat("Weight not in range", "\n")}
}
```

# Part 4: applying

Oftentimes, we will want to apply a function to every row or every column of a
data frame, a matrix, or a list. R includes functions to do this with and
without resorting to `for` loops.

`MouseBalanceTimeSeries` is a matrix where two strains of mice (B6 - alcohol
preferring and D2 - non-preferring) are given alcohol followed by a balance beam
test in triplicate (measured in seconds). They are then treated with an
experimental treatment and then retested on the balance beam after 1 hour.

Note that the mouse identifiers (rownames) in `MouseBalanceTimeSeries` are
identical to those in `MouseFrame`.

```{r}
rownames(MouseFrame) == rownames(MouseBalanceTimeSeries)
MouseBalanceTimeSeries
```

**QUESTION 4-1**: What do you notice about this time series? Is the data
complete?
```{r}
# space for your answer here

```

Let's compare row means (within-mouse). We can do this using `apply()`: The
margin argument sets whether we apply the function to rows (`MARGIN=1`) or
columns (`MARGIN=2`).

```{r}
apply(X=MouseBalanceTimeSeries[1:4,], MARGIN=1, FUN=mean)
```

**QUESTION 4-2**: What is the output of the above command? Is it a matrix or a
vector?
```{r}
# space for your answer here

```

**QUESTION 4-3**: Compare the row means of the first three columns
(pre-treatment) to the last three columns (post-treatment) for the first four
rows of `MouseBalanceTimeSeries`. Would you say there is a difference between
pre- and post-treatment mice?
```{r}
# space for your answer here

```

You may notice that there are some missing values (`NA`s) in the dataset. There
are multiple ways to dealing with missing values. Ideally, we want to knkow why
the values are missing. In this case, the values are missing because the mice
were not available for the balace beam experiment.

`na.omit()` works with data frames and matrices, and will eliminate rows where
values are missing. Sometimes this is what you want, sometimes is isn't. Many
times, your dataset is going to have a small number of subjects and it may be
possible to achieve your objective with missing data. 

Another method that may be useful is interpolation of missing values (i.e., for
time series data).

```{r}
MBTSfiltered <- na.omit(MouseBalanceTimeSeries)
```

We can see the rows filtered out as an attribute called `na.action` on the
MBTSfiltered object:

```{r}
attributes(MBTSfiltered)$na.action
```

## Applying over a set of factors

`tapply()` is for when you have a vector for grouping (such as a factor) and you
want to apply a function to each group. Here we take the mean weight by Strain.

```{r}
tapply(X=MouseFrame$Weight, INDEX=MouseFrame$Strain, FUN=mean)
```

**QUESTION 4-4**: Use `tapply` to compute the first (pre-) and fourth (post)
column means by `Strain` (remember, the identifiers are identical in both
`MouseFrame' and `MouseBalanceTimeSeries`, so you can use the `Strain` column in
`MouseFrame` on `MouseBalanceTimeSeries`. Remember to remove missing values.
Does the data appear to have `Strain` specific differences in column meants? If
so, between which comparison? (B6 pre- to B6 post-, or B6 pre- to D2 pre- or
...)
```{r}
# space for your answer here

```

There are many other `apply`-type methods. One of the best packages for
aggregation/apply methods is Hadley Wickam's `plyr` package. It is a bit quirky,
but provides efficient ways to apply functions across a variety of data
structures.

`sapply()` and `lapply()` are both methods to work on list-like objects (such as
vectors and lists). `lapply()` works strictly on lists of objects, which we'll
talk about in the next section. `sapply()` is much more flexible in terms of
output.

`mapply()` is used when you have a function that takes multiple arguments.

# Part 5: Lists, Batch processing, and Storing Results

Lists are a general data structure. Each slow can be named and can hold any R
Object.

```{r}
data(iris)
testList <- list(a=c(1,4,5), b=iris, c="STRING")
```

Accessing the actual element of the list can be done by name or by number. Note
the use of the double brackets `[[]]` to access the element.

```{r}
testList[[1]]
testList[["a"]]
```

Another way to access list elements is using `$` (note this only works with
named elements).

```{r}
testList$a
```

We can also assign objects using these accessors.

```{r}
testList[["d"]] <- "ANOTHER STRING"
```

Confusingly, we can also access elements within a list element. The way to think
about this i that the double backets are accessing the object, and then we can
subset within the object.

**QUESTION 5-1**: Guess what each of the following statements accesses from
`testList`. Was your guess correct?
```{r}
testList[[1]][2]
testList[["b"]][1:5,]
testList$c
dd <- "a"; testList[[dd]]
```

Note that when accessing the double brackets, the difference between "a" (a
string) and `dd` (a variable which has "a" stored in it). 

By itself, a list seems somewhat useless. However, a really good use for them is
to store results when you're batch processing (i.e. processing a bunch of files.

Let's load three different sets of data (from three difference experiments).
each file beginning with `mouseExpt`, so let's use that as a pattern we'll store
the results in a list.

```{r}
mouseExptList <- list()
mouseExps <- list.files(pattern="mouseExpt")
for(mm in mouseExps){
    mdata <- read.table(mm, sep="\t", header=TRUE, row.names=1)
    mouseExptList[[mm]] <- mdata
}
```

**FINAL PROBLEM**: For each data frame in `mouseExptList`, calculate the row
sums, and add it as a column to the data frame with the name `rowMean`. Save
each moditied data frame into a new list with the same names for each slot. 

Do you notice anything about the data? What would be one way to deal with the
issues in the data?
```{r}
# space for your answer here

```
