---
title: "Module 2: Vectors, Data Frames, Subsetting, Filtering, and Ordering"
author: "Ted Laderas"
date: "July 9, 2015"
output: html_document
---

## Overview

1. Vectors
2. More about Data Frames
3. Subsetting
4. Filtering
5. Sampling and Ordering

## Setup Workspace

Before you start: remember to set your working directory to the `module2` folder!

Now, let's load up the workspace:
```{r}
load("module2.RData")
```

## Part 1: Vectors

Let's look at a vector in the workspace.

```{r}
weights
```

Note that each slot in a vector can have a unique name.

```{r}
names(weights)
```

Length is one of the properties a vector has.

```{r}
# A vector has a length
length(weights)
```

The `weights` vector can be subsetted. Let's get the very first weight.

```{r}
weights[1]
```

If we pass a sequence to the vector, we can subset. Here we want to grab the
first 5 weights, so we pass a sequence from 1 to 5.

```{r}
# A sequence is actually a vector
1:5

# Get the first 5 weights
weights[1:5]
```

Values in a vector can also be retrieved by name:

```{r}
weights["M15"]
```

**QUESTION 1-1**: How would we get the last 5 weights?
```{r}
# Space for your answer here
```

**QUESTION 1-2**: How would we get the first 5 weights in reverse order?
```{r}
# Space for your answer here
```

### Assignment in Vectors

In general, assignment is done by using the `<-` operator. Here we assign the
first 5 elements of the weights vector to `b`.

```{r}
b <- weights[1:5]
b
```

### Operations on Vectors

In general, if you can do an operation on a vector all at once, try to do it (as R is optimized for this).

```{r}
# Run the mean on a vector
meanWeight <- mean(weights)
```

Missing numeric values tend to be coded as `NA`s. Not all functions will handle
`NA`s the same. In module 3, we will talk about the `na.omit()` function.

### Vector Operations
Some operations output a vector of the same length. Let's get the residual sum
of squares.

```{r}
resids <- weights - meanWeight
```

**QUESTION 1-3**: What happened here? Why can we subtract a scalar from a vector?
```{r}
# Space for your answer here
```

**QUESTION 1-4**: What happens when you try to subtract two vectors of unequal
lengths? I.e. `weights[1:5] - weights[1:2]`
```{r}
# Space for your answer here
```

Here's another example of vector operations:
```{r}
# Look at resids
resids

# Square the residuals
sqResids <- resids^2

# Produce residual sum of squares
sum(sqResids)
```

#### Making vectors
In general, you can use the `c()` (concatenate) command to make vectors.

```{r}
vec <- c(1,2,3:5)
vec
```

You can also concatenate a vector to another vector.

```{r}
vec2 <- c(vec,5,5)
vec2
```

**QUESTION 1-5**: What happens when you try to mix characters and numbers? What
does this tell you about vectors? I.e. `c("This is character data", 1:5)`
```{r}
# Space for your answer here
```

**QUESTION 1-6**: How might you combine `c()` and a series of names to pull out
only the "M1", "M5", and "M10" weights? This is another kind of subsetting
operation, and we'll cover it more in detail later.
```{r}
# Space for your answer here
```

### Descriptive Statistics and Exploratory Data Analysis

You can run operations on a single vector.
```{r}
# Show mean and quantiles
summary(weights)

# Show a histogram
hist(weights)
```

**QUESTION 1-7**: What can you say about the distribution of weights? i.e., are
they distributed normally?
```{r}
# Space for your answer here
```

### Vector Datatypes and Casting

There are four useful vector datatypes to keep in mind: 

- numeric, 
- character,
- boolean, and
- factor.

#### Numeric

We've already seen numeric vectors:
```{r}
c(1,2,3,5)
```

Another useful way to initialize a vector is by using a sequence:
```{r}
c(1:10)
```

#### Character

Characters can also be put into vectors:
```{r}
c("foo", "bar", "baz")
```

#### Boolean

Boolean vectors are simply `TRUE`/`FALSE`:
```{r}
c(TRUE,TRUE,TRUE,FALSE)
```

To some extent, we can convert between vector types using `as.numeric()` and
`as.character()`.
```{r}
as.character(c(1,2,3,5))
```

#### Factors

Factor vectors can be ordered (by specifying the level argument) or unordered
(by not specifying the level argument). The ordering impacts properties such as
the order the factors are plotted, the order in which factors are reported in
table(), and the way other functions treat them. 

Note that levels have to contain all factors.
```{r}
testVecFac <- factor(c("D", "D", "E"), levels=c("D", "E"))
```

Character and factor vectors are for the most part interchangable for tables.
```{r}
testVecChar <- c("A", "B", "C","C", "D")
testVecFac <- factor(c("A", "B", "C","C", "D"))

table(testVecChar)
table(testVecFac)
```

However, when we use numbers as factors, be aware that `as.numeric()` doesn't
work. This is due to the internal representation of factors as integers.
```{r}
testVecFac <- factor(c(1,5,5,56), levels=c(1,5,56))
as.numeric(testVecFac)
```

Instead, we must first cast the vector to character, and then numeric:
```{r}
as.numeric(as.character(testVecFac))
```

A final note: you may notice that `read.table()` will treat any strings as
factors when loading. You can override this behavior by setting the argument
`stringsAsFactors` to `FALSE` (refer to `?read.table` for more information.)
whether you will want your strings represented as characters or factors is
dependent on the application.  

```{r}
testTable <- read.table("mouseData.txt", header=TRUE)
summary(testTable)

# You can always cast a factor as a string, remember
testTable$Strain <- as.character(testTable$Strain)
summary(testTable)
```

Let's try to read the table as characters.
```{r}
testTable2 <- read.table("mouseData.txt", stringsAsFactors=FALSE, header=TRUE)

# Note the numeric property of `Weight` is preserved.
summary(testTable2)
```

#################################
###Part 2: More about Data Frames
#################################
##data.frames consist of associated vectors that have the same length
##in a matrix-like format
##columns do not have to be the same data type; can mix
##boolean, character, factor, and numeric data types in the same data frame

##show a dataframe example (that we have already seen)
MouseFrame

##dataframes have two properties: number of rows
##and number of columns

nrow(MouseFrame)
ncol(MouseFrame)

##can also get both of these properties by dim
dim(MouseFrame)

##show only the first five rows
##useful for checking that you read in data properly
MouseFrame[1:5,]

##QUESTION: note that the column of each data frame is pretty much a vector, but a
##row is actually a list. Why is that?

```{r}
#space for your answer here
```


##data frames have two kinds of names associated with them:
##column names (usually corresponds to variables):
colnames(MouseFrame)

##row names (usually subjects (such as patients) or gene names)
##one requirement of rownames is that each row name is unique
##so that subsetting by name will return a unique result. Patients
##and genes tend to have a unique identifier that we can use. In
##this case, the unique identifier is the MouseID
rownames(MouseFrame)

##access only the Weights column of MouseFrame
MouseFrame$Weight

####QUESTION 2-1: what happens when you access the Gender column?
####How does it differ from the weights column?
####Hint: use class() to compare the different columns

##summarize mouse frame
summary(MouseFrame)

####QUESTION 2-2: How many Males and Females are there?
####Is this a balanced design?

##initializing a dataframe from a set of vectors
factorVec <- factor(c("A", "A", "B", "C", "C"))
massVec <- c(12, 1.4, 2.4, 6.4, 10)
nameVec <- c("Moe", "Minnie", "Curly", "Mickey", "Larry")

dF <- data.frame(nameVec, massVec,factorVec)

##note that column names are derived from the original vector names
##however, using assignment, you can change the column names when you 
##initialize them.
##in terms of reproducibility, try to make a new copy of a data frame
##whenever you manipulate the contents.  This leaves a clear audit trail
##in your script.

dF2 <- data.frame(idVec=nameVec, mass=massVec,class=factorVec)

##can also add an extra column to a data frame

anotherVec <- c(1,4,4,10,10)

dF3 <- data.frame(dF2, anotherVec)

##accessing individual columns can also be done using $
##this is useful for filtering operations

dF3$anotherVec

##if you need to do a lot of column based opertions on a data.frame,
##you can use attach()
attach(dF3)
idVec
mass

##when you're done with manipulating the data frame, remember to detach it.
##otherwise things get confusing, especially when you have data frames
##with identical column names.

detach(dF3)


#######################################################
##Part 3: Subsetting
#######################################################
##Many data structures can be subsetted to produce a smaller result set
##We've encountered this by using numerics, grabbing the first five rows
##mouseFrame
MouseFrame[1:5,]

##we can mix subsetting vectors as well, grabbing only the first five rows and
##the Strain and Weight Columns
MouseFrame[1:5,c("Strain","Weight")]

##if we don't want a column or row, we can remove it by using the "-" operator.
##note that this doesn't work for any index vectors except the numerical index.
##I tend not to use this and instead will use a colnames or rownames subset as above.
MouseFrame[-c(1:3), -1]

##another way of subsetting is by using a boolean vector, which seems kind of dumb,
##until you realize that you can generate this boolean vector using other methods.
##(see Part 4).  In general, we are trying to avoid hardcoding values, and subsetting
##on specific criteria makes our code more generalizable.

MouseFrame[,c(TRUE,TRUE,FALSE)]

####QUESTION 3-1: what happens when you use a boolean vector that's shorter than the number
####of columns? What happens with the remaining columns?  Are they returned?

```{r}
#space for your answer here
```


####QUESTION 3-2: How would we select every third row? Hint: help("%%")

```{r}
#space for your answer here
```


#######################################################
##Part 4: Filtering
#######################################################
##filtering operations are done by generating a boolean vector using a comparator
##such as <, >, or ==
##Here's how to filter on a vector to select those weights with only 

weights[weights < 50]

####QUESTION 4-1: What does the comparator operator 
####weights < 50
####generate? Why does it work?

#Filtering on a data frame is identical to filtering on a vector
#note that we must filter on a column as a criterion.
#specifically, both MouseFrame$Gender and 
#subsetting data frames - filtering
#select only females

MouseFrame[MouseFrame$Gender == "F",]

####QUESTION 4-2: What does 
####MouseFrame$Gender == "F" 
####do?  Why does it select the correct cases?

```{r}
#space for your answer here
```

####QUESTION 4-3: how do you select females of strain D2?
#Hint: it's ok to use more than one operation
#Hint: you can also chain selections using the & (and) or the | (or) operators
#Hint: don't forget the comma!

```{r}
#space for your answer here
```

####QUESTION 4-4: How would we select those mice with a weight
####less than 50 grams?

```{r}
#space for your answer here
```


##chaining subsetting and filtering operations
##the difference between subsetting and filtering is largely semantic, but you
##can think of filtering as part of subsetting operations using some sort of
##conditional operators.

##because they are essentially producing the same thing (boolean vectors), you can 
##use a combination of subsetting or filtering operators to produce a reduced dataset

MouseFrame[MouseFrame$Strain == "B6", c("Strain","Weight")]

##many statistical methods may require you to recode a group
##ifelse() works very well for this when you have two groups. 
##ifelse() takes a test condition, such as a filtering operation, and
##will produce a code (in this case 1) if the condition is true and another
##code (in this case 0) if the condition is false.

lowWeight <- ifelse(test = MouseFrame$Weight < 45, yes = 1, no = 0)

####QUESTION 4-5: produce a cross table between MouseFrame$Strain and lowWeight
####(you may need to cast lowWeight as a factor).  Do Weight and Strain appear to
####be associated?

```{r}
#space for your answer here
```

##############################
#Part 5: Sampling and ordering
##############################
###Both sampling and ordering can be thought of as subsetting operations as well.
###Let's sort MouseFrame by Weight

####QUESTION 5-1: what does order() do? How can you use it to sort a data frame?
order(MouseFrame$Weight, decreasing=TRUE)

```{r}
#space for your answer here
```

###we can also specify sorting by multiple factors
order(MouseFrame$Weight, MouseFrame$Gender)

####QUESTION 5-2: How can we sort the data frame by Strain then Weight?

```{r}
#space for your answer here
```


#Let's produce bootstrap samples of our dataframe

####QUESTION 5-3: What does the sample() code below do?  How
#####can we use it to produce samples of our dataframe?
sample(1:nrow(MouseFrame), 5, replace=TRUE)

```{r}
#space for your answer here
```


####QUESTION 5-4: Produce a bootstrap sample without replacement of 7 rows
####of MouseFrame. Confirm that you have sampled unique rows in your dataframe.

```{r}
#space for your answer here
```


###A note about make.names()

###***FINAL PROBLEM: Load the iris dataset using data(iris). (this will load the data
###***as the iris object).  
###***
###***Produce two different data frames for the setosa and virginica
###***species, selecting the Petal.Length and Petal.Width columns.  
###***
###***How many setosa samples have a Petal Length less than 3.5? 
###***How many virginica samples have a Petal Width less than 2.0?
