---
title: "Module 4: Merging, Visualizing, Functions"
author: "Ted Laderas"
date: "August 22, 2015"
output: html_document
---

Remember: change your working directory.

Load the module workspace.

```{r}
load("module4.RData")
# look around
ls()
```

## Part 1: Merging

### Merging dataframes

Suppose you have two dataframes that have a common identifier (such as MouseID)
and you want to merge them. If you want to merge on row names, you can use the
following form: `merge(MouseFrame, MouseCovariates)`

If you want to merge on an identical column names in both data frames, this works:
`merge(MouseFrame, MouseCovariates, by="MouseID")`

If the column names were different, we can specify them by using the by.x and
by.y arguments: 
`merge(MouseFrame, MouseCovariates, by.x="MouseID", by.y="MouseID")`

These merges are known as **natural joins**, because we match every element that we
can in the first frame with the second. If we insist on including every row in
the first frame, even though it doesn't occur in the second frame, it's called a
left outer join.  Think of the first frame as the left frame, and the second
frame as the right frame.

```{r}
merge(MouseFrame, MouseCovariates, by = "MouseID", all.x =TRUE)
```

Similarly, if we insist on including all the rows in the second frame even though
they don't occur in the first, it's called a **right outer join**.

```{r}
merge(MouseFrame, MouseCovariates, by = "MouseID", all.y =TRUE)
```

**QUESTION 1-1** How would we specify a full outer join, i.e., a join that includes
all rows from both frames, regardless of match?
```{r}
# Space for your answer here
```

Note that these kinds of joins are also used in Databases, which we will talk a
little about in the next module.

**QUESTION 1-2** Read the documentation for `?merge`. Do a natural join of
`MouseFrame` and `MouseBalanceTimeSeries` and do not sort by `MouseID`.
```{r}
# Space for your answer here
```

##############################
###Part 2: Visualizing Data
##############################

##We've alreadly seen some methods for plotting 1-dimensional data, such as 
##hist() and boxplot(). boxplot() will also work on data.frames and give you
##a boxplot per variable.

##we need to cast MouseBalanceTimeSeries as a data.frame since it is a matrix:
boxplot(data.frame(MouseBalanceTimeSeries), main="Before and After")

##We can label the x axes and y axes by supply the xlab and ylab argument
boxplot(data.frame(MouseBalanceTimeSeries), main="Before and After Treatment",
        xlab = "Condition", ylab = "Time on Balance Beam (s)")

##This almost looks good, but we can't read the labels for the different
##experiments. If we read the help file for ?par() (the graphical parameters),
##we can adjust those labels by specifying a las = 3 argument:

boxplot(data.frame(MouseBalanceTimeSeries), main="Before and After Treatment",
        xlab = "Condition", ylab = "Time on Balance Beam (s)", las=3)

##let's remove the x label to make it clearer:

boxplot(data.frame(MouseBalanceTimeSeries), main="Before and After Treatment",
        ylab = "Time on Balance Beam (s)", las=3)

##what about xy plots? R is equipped to do those as well.
data(iris)
plot(iris[,1], iris[,2])

####QUESTION 2-1. Add arguments to the above plot. Label the x-axis "Sepal Length"
####and the y-axis "Sepal.Width".  Make the title "Iris Dataset". Save it to a pdf
####with the name "Iris-plot.pdf".

####QUESTION 2-2. Read the documentation for pairs(). Try it out on the iris dataset
####(take out the "species" variable.)
####What does this plot do? Why is it useful?

##what about pie charts? Pie charts are evil. They are a poor representation of data,
##and make proportions difficult to compare.  If you really must, read the help file
##for ?pie.  

##as an alternative, consider the stacked bar plot.  
##subset first ten rows so barplots aren't equal sized
smFrame <- MouseFrame[1:10,]
##tabulate the Gender/Strain cases
tab <- table(smFrame$Gender, smFrame$Strain)
##produce a stacked barplot allowing to compare the distribution of males/females
##within B6 and D2 strains
barplot(tab, col=c("blue", "red"), legend=rownames(tab), main = 
    "Barplot comparing Gender within Strain")

##############################
###Part 3: Functions and Scope
##############################
##Let's make a function.  Functions are defined with the command function() and 
##a signature, which describes the arguments.

sqrFun <- function(x) {
  sqrVal <- x^2
  #return the value
  sqrVal
}

##note that we didn't define what x has to be. For S3 functions, there is no 
##explicit type checking. S4 objects and methods (which are beyond this bootcamp),
##do have explicit signature checks, which can be helpful in software development.
##For right now, however, don't worry about them.

####QUESTION 3-1: Predict the output of each of these invocations of sqrFun()
sqrFun(5)
sqrFun(c(1,2,3))
sqrFun("STRING")

##Using identity functions such as class() as is.data.frame()
##or is.matrix(), we can be more explicit about our inputs.

sqrFun <- function(x) {
  if(is.numeric(x)){
  sqrVal <- x^2
  #return the value
  sqrVal}
  else{stop("Not a numeric input")}
}

####QUESTION 3-2: Change the above function to accept only single numeric inputs.
####Hint: use length()

##If you'd like to return multiple values, such as mean and sd, you can use a list.
##it's best to use a named list to return results.

meanSdFun <- function(vec){
  #test to see if input is a vector and at least length 3
  if(is.numeric(vec) && length(vec) > 2){
      meanRes <- mean(vec)
      sdRes <- sd(vec)
      #store the results in a list
      resList <- list(mean=meanRes, sd=sdRes)
      #return results
      resList
  }
  else {stop("Not a numeric vector")}
}

results <- meanSdFun(c(2,5,7))
results[["mean"]]
results$sd

##We can also define default values for arguments by specifying them in the signature:

plotXYFunc <- function(x,y,linecolor="red", plotTitle="X vs Y"){
  plot(x,y,col=linecolor, type="l", main=plotTitle)
}

##use default arguments
plotXYFunc(x=c(1,2,3),y=c(2,4,6))
##specify arguments
plotXYFunc(x=c(1,2,3),y=c(2,4,6), linecolor="blue", plotTitle="Stuff vs Stuff")


####QUESTION 3-3: Modify plotXYFunc to include a "fileout" argument with default
####value "XYplot.pdf". Change the line type to stairsteps (refer to 
####http://www.statmethods.net/graphs/line.html for more info). Modify the code to 
####output a pdf file with fileout as its name.  If fileout=NULL, simply display 
####the graph instead of writing it to a file.


##One important thing we need to discuss about functions. That of variable scope.
##Variable scope has to do with whether a variable "exists" in a certain "universe".
##This 'universe' could be the workspace (global scope), or a function (local scope).  
##We've mostly been using global variables by defining them in the workspace.
##
##However, in functions, we can also define variables on the fly.  However, because
##of scope, you can't access a variable that was defined in a function in the workspace.
##for example, say I make a function and define an internal Variable intVar:
##
##another way to think of it, is if you define a variable in brackets, anything in those
##brackets can use that variable. Outside of those brackets, the variable doesn't exist.

testFun <- function(xVec, yVec){
  intVar <- c(xVec, yVec)
  outVar <- intVar^2
  outVar
}

testFun(xVec=c(1,5),yVec=c(4,6))
##intVar doesn't exist outside of the function (local scope)
intVar

##however, if we globally define a variable in the workspace, a function can "see" this
##variable and access it.  

globVar <- 30

testFun2 <- function(x){
  out <- x + globVar
  out
}

testFun2(50)

##This is the cause of many headaches, in that you can write a function that depends on 
##a global variable, attempt to run the function in a workspace where that global 
##variable isn't defined, and get an error.
##
##The point is, be careful about naming variables and be explicit about passing variables
##into functions. It's easy to prototype code that accidentally depends on global 
##variables that are not explicitly passed into the function. 

#Have you wondered what the hell the "..." is in function descriptions? Basically when 
#a function has "...", it means you can pass it additional arguments to internal functions
#This is really useful when using things like apply and you have a function that requires
#additional arguments.

###***FINAL QUESTION: Write a function with the following signature:
###***meanWithNA <- function(x)
###***Assume x is a vector and has NAs. Write the function such that it eliminates NAs
###***and returns the mean of the remaining values in the vector. If the length of 
###***remaining values is 0, have the function return NA.

###***Use this function to revise your rowMeans for each data.frame in mouseExpList 
###***(provided as an data object in module4). 
###***Write another function with the following signature:
###***compareMTwithRowMean <- function(mazeTime, rowMean)
###***and have it output a vector that is 1 if mazeTime < rowMean, 0 if not. 
###***Apply this function to mazeTime3 and rowMean. Count the number of 1s and 0s
###***for each data set.  Based on this number, do you think the mice learned how
###***to run the maze faster?



