---
title: "Module 5: Packages, Intro to Databases, and String Manipulation"
author: "Ted Laderas"
date: "June 24, 2016"
output: html_document
---

#Getting Started

Before you get started, check out `database-introduction.pdf` in this folder if you are not familar with basic concepts of databases.

Remember to set your working directory for this current folder!

#Part 1: Installing packages

Most of the functionality of R has been extended by other programmers through the package system.  There are currently two sets of repositories to get packages.  The first is the CRAN (Comprehensive R Network) system, which is accessed through the install.packages() command.  These packages tend to be statistical packages that do not have necessarily a biological focus.

The second package repository is Bioconductor, which focuses on biological applications. Installation of these packages require you to first install Bioconductor's base distribution. We'll talk more about this in part 3.

Let's install the database package that we need. Follow the prompts and select a mirror nearby you to download if prompted. RStudio has a default mirror so it may not prompt you.

```{r eval=FALSE}
install.packages("RSQLite")
```

To load the package into memory, you can use `library()` or `require()`. Note that you don't need quotes to access the package.

```{r}
library(RSQLite)
```

To see what documentation exists for a package, you can use:

```{r}
help(package="RSQLite")
```

Some packages may also contain Vignettes, which are short documents that usually show you how to use a package. You can see what vignettes are available by using

```{r}
vignette()
```

Depending on the package, you may have noticed that some extra packages haveeither loaded (if they were already installed) or were downloaded (if you didn't have them already) These are packages that RSQLite depends on (or dependencies). For example, when you load RSQLite, the DBI package is also loaded.

#Part 2: Databases

If you haven't yet loaded the RSQLite package, do it now. We'll explore a simple database.

```{r}
library(RSQLite)
```

Let's initialize a database connection object. This object gives us access to the database. Note that our database is a local SQLite database, but we can instantiate database connection objects for databases that exist on other servers in the internet.

```{r}
dbConn <- dbConnect(SQLite(), "module5.sqlite")
```

The first thing we want to do is figure out what the structure of the database is. That is, we want to know what the names of the tables and the names of the fields are.

```{r}
dbListTables(conn=dbConn)
```

Next, we can list all the fields within a specific table:

```{r}
dbListFields(conn=dbConn, name="Assignment")
```

To send queries to our database, we use the `dbSendQuery()` command. Note that this command will execute our query, and return a result set object, but not actually a data frame. To get the actual data, we need to use `fetch()`.

```{r}
SQLquery1 <- "SELECT * FROM Assignment"
SQLresult1 <- dbSendQuery(conn=dbConn,statement = SQLquery1)
```

`fetch()` includes an "n" argument which is the number of rows you'd like returned. To return everything, use `n = -1`. This is a toy example, so we can load all the results into memory.  You are probably going to be utilizing *very* large databases whose tables can't fit into memory, so we'll also show you how to iterate with `fetch()` a little later.

```{r}
AssignmentFrame <- fetch(SQLresult1, n=-1)
head(AssignmentFrame)
```

To free up memory and other resources, remember to clear the result afterwards.

```{r}
dbClearResult(SQLresult1)
```

For larger memory tables, we can use a while loop to fetch the results a few rows at a time.  We can use a `while` loop and the `dbHasCompleted()` test to fetch these results.

```{r}
SQLquery1 <- "SELECT * FROM Assignment"
SQLresult1 <- dbSendQuery(conn=dbConn,statement = SQLquery1)
```

`dbHasCompleted()` returns a boolean whether the fetch has retrieved all rows (TRUE) or not (FALSE) in your result object. In this case, we haven't fetched any results yet, so the output should be FALSE.

```{r}
dbHasCompleted(SQLresult1)
```

Let's retrieve some rows a few rows at a time.

```{r}
##fetch the first row to initialize the data frame
resIterateFrame <- fetch(SQLresult1, 1)
##fetch rows until dbHasCompleted(SQLresult) == TRUE
while(dbHasCompleted(SQLresult1) == FALSE){
  #we can rbind our fetch results directly to resIterateFrame 10 rows at a time
  resIterateFrame <- rbind(resIterateFrame, fetch(SQLresult1, 10))
}

head(resIterateFrame)
```

Don't forget to clear the results!

```{r}
dbClearResult(SQLresult1)
```

Let's  confirm that the results are identical in number of rows fetched

```{r}
nrow(AssignmentFrame)
nrow(resIterateFrame)
```

**PROBLEM 2-1**: Fetch the Mouse table from the database. Compare the number of rows of the Assignment Table to the number of rows of the Mouse table.  Are the numbers what you expected?

```{r}
#Space for your answer here
```

Similarly, we can filter our results using the WHERE statements. Note how similar this is to filtering a data frame. You can use all your data frame skills here!

```{r}
SQLquery2 <- "SELECT * from Mouse WHERE StrainID = 'B6'"
SQLresult2 <- dbSendQuery(conn=dbConn,statement = SQLquery2)
mouseQuery <- fetch(SQLresult2, n=-1)
head(mouseQuery)

dbClearResult(SQLresult2)
```

**PROBLEM 2-2**: Select all mice that have a weight less than 60 grams (You'll have to use a comparator such as "<"). Not counting the NAs, how many mice are there?

Let's start doing joins.  We know from our Diagram in `database-introduction.pdf` that ExperimentID in Assignment maps to ExperimentID in Experiment, so let's join the tables together using a "WHERE" query:

```{r}
SQLquery3 <- "SELECT * FROM Assignment, Experiment WHERE 
  Assignment.ExperimentID = Experiment.ExperimentID"
SQLresult3 <- dbSendQuery(dbConn, SQLquery3)
resFrame3 <- fetch(SQLresult3, n=-1)
head(resFrame3)

dbClearResult(SQLresult3)
```

If we are interested in only specific fields, we can specify them as part of the SELECT statement before the FROM clause, using a "." notation to specify the table and field:

```{r}
SQLquery4 <- "SELECT Experiment.Date, Assignment.LabtechID FROM
            Assignment, Experiment WHERE Assignment.ExperimentID = Experiment.ExperimentID"
SQLresult4 <- dbSendQuery(dbConn, SQLquery4)
resFrame4 <- fetch(SQLresult4, n=-1)
head(resFrame4)

dbClearResult(SQLresult4)
```

Let's see if LabtechId and Date are correlated (i.e., whether the two labtechs worked experiments on separate dates):

```{r}
table(resFrame4$Date, resFrame4$LabtechID)
```

**PROBLEM 2-3**: Join the Mouse and Assignment Tables on MouseID and fetch the results. Compare the number of rows in the query result to the number of rows in the Mouse Table. What is your conclusion?

```{r}
##Space for your answer here
```

Databases show their power when we add multiple conditions to the WHERE clause, using AND:

```{r}
SQLquery5 <- "SELECT * from Mouse WHERE StrainID = 'D2' AND Weight < 70"
SQLresult5 <- dbSendQuery(dbConn, SQLquery5)
resFrame5 <- fetch(SQLresult5, n=-1)
head(resFrame5)

dbClearResult(SQLresult5)
```

Much like our work with the merge() function in Module 4, we can do Left outer joins. The syntax is a little different in that we have to specify which is the left table and which is the right table by specifying `(Left Table) LEFT JOIN (Right Table) ON`

Unfortunately, right joins and full joins are not implemented in SQLite, so you have to do the majority of the work with the left outer join.

```{r}
SQLquery6 <- "SELECT * FROM Mouse, Assignment LEFT JOIN Assignment ON Assignment.MouseID = Mouse.MouseID"
SQLresult6 <- dbSendQuery(dbConn, SQLquery6)
resFrame6 <- fetch(SQLresult6, n=-1)
head(resFrame6)

dbClearResult(SQLresult6)
```

Left table results that don't map to the right table will show columns with NAs in the right table columns.  Note that in this result, we return all of the columns for both tables, so there are two columns called MouseID.

**PROBLEM 2-4**: How could we find all the mice that don't have experiment assignments?
Hint: use your `data.frame` skills and `is.na()`.

```{r}
###Space for your answer here
```
  
Let's do a query joining all of the tables.

```{r}
SQLquery7 <- "SELECT * FROM Mouse, Experiment, Assignment WHERE Mouse.MouseID = 
  Assignment.MouseID AND Experiment.ExperimentID = Assignment.ExperimentID"
SQLresult7 <- dbSendQuery(dbConn, SQLquery7)
resFrame7 <- fetch(SQLresult7, n=-1)
head(resFrame7)

dbClearResult(SQLresult7)
```

Writing Tables can be achieved with a dbWriteTable() command. Let's save our new table in the database:

```{r eval=FALSE}
dbWriteTable(dbConn, name="allJoinResults", resFrame7)
```

Important note. When you are done, to disconnect from the database, use `dbDisconnect()`. This is necessary to save the tables. 

```{r}
dbDisconnect(conn=dbConn)
```


##Further reading: 

For a full list of SQL commands that SQLite understands check this link out: http://www.sqlite.org/lang.html

##FINAL PROBLEM 

Join all three tables using the associated IDs (`MouseID`, `ExperimentID`) and look at the following:

1) Is there a Strain-specific difference in the mean Balance Time? (Use your tapply() skills!)

2) Are there any other categorical variables associated with strain in the data? (Make some cross tables like I did with Date and LabtechID)

3) If so, does this make your conclusion from 1) stronger or weaker? Why?

```{r}
##Space for your answer
```

##Part 3: String Manipulation

The following section is optional, but you might find it very helpful.

First of all, obviously, a string can be initialized with ""

```{r}
seqExample <- "AATTGGTTCCTT"
```

Strings have some useful properties. For example, number of characters:

```{r}
#number of characters in a string
nchar(seqExample)
```

```{r}
##nchar() == 0 is the only way to test for an empty string
stringVec <- c("Test", "Test2", "", "Test3")

##let's define a function testEmptyString that tests for an empty string
##and returns NA if empty, the value, otherwise

testEmptyString <- function(x){
  if(nchar(x) != 0){return(x)}
  else{return(NA)}
}

#try it out:
testEmptyString("")
testEmptyString("String")

#let's apply it to stringVec
sapply(stringVec, FUN=testEmptyString)

#substr can be used to extract substrings
#usage: substr(string, start, end)
#where start and end are the locations in the substring 
#you want to extract
#should extract "GGTTCC"
substr(seqExample, 5, 10)

####QUESTION 3-1: print each letter separately from the seqExample above.

####QUESTION 3-2: count the frequency of each letter in seqExample.
####Hint: you can store the counts in a vector

#More string tricks
#strsplit can split a string by a separator into a vector
#note that returns a list, so you will need to use the [[]] annotation
#to access the actual vector
#This is especially useful when there are multiple delimiters in a text file.
test <- "It,Slices,And,Dices"
spl <- strsplit(test,split=",")
spl[1]
spl[[1]]

#strsplit actually is made to work on a vector of characters
test2 <- c("It,Slices,And,Dices", "And,Costs,19.99")
strsplit(test2, ",")

####QUESTION 3-3: What happens when you specify a "" separator?

#paste can glue strings back together into a single string
paste("The", "Quick", "Brown", "Fox", "Etc.", sep=" ")

####QUESTION 3-4: What happens here?
vec <- c("The", "Quick", "Brown", "Fox", "Etc.")
paste(vec, "-ay", sep="")

#grep() can be very useful to search for a string within a character vector
#it returns the indices of all matches
grep("is", c("This","is", "my", "Thesis", "Statement"))

#We don't have time to cover regular expressions, but you can use them in R
#The regular expressions are in Perl syntax.
```
